---
title: "01 Metadata Preparation"
editor_options: 
  chunk_output_type: console
---

## Define the input and output paths

```{r}
# input
metadat_file <- here::here("data", "raw", "2023_05_29_clean_metadata_LIMS.csv")
phy_file <- here::here("data", "raw", "Metaphlan3Counts_psObject.rds")

# output
out_dir <- here::here("data", "processed")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
```

## Set-up

```{r}
# load libraries
library(magrittr)
suppressPackageStartupMessages(library(tidySingleCellExperiment))
```

## Metadata Preprocessing

### Load & Inspect Metadata

```{r}
# Load
meta_df <- readr::read_csv(metadat_file, show_col_types = FALSE)

# Inspect
skimr::skim(meta_df)
```

### Clean Metadata

- Remove columns with unit information
- Remove columns with results dates
- Remove columns with only one unique value
- Remove columns with more than 50% missing values
- Remove `<` and `>` from sCD14 and transform to numeric
- Rename levels of group variable and rename as tretament
- Create time_point_n (nueric) variable from time_point (factor)

```{r}
# Clean
meta_df <- 
  meta_df %>% 
  dplyr::select(-dplyr::starts_with("units_")) %>%
  dplyr::select(-dplyr::starts_with("result_date_")) %>%
  dplyr::select_if(~ length(unique(.x)) > 1) %>%
  dplyr::select_if(~ mean(is.na(.x)) < 0.55) %>% 
  dplyr::mutate(
    sCD14 = as.numeric(stringr::str_remove(sCD14, "<|>")), 
    time_point_n = time_point,
    time_point = factor(time_point),
    treatment = dplyr::if_else(group == "bDRV", "DRV/r", group)
  )
```

### Save Cleaned Metadata

```{r}
readr::write_csv(meta_df, here::here(out_dir, "metadata_cleaned.csv"))
```

## Prepare Phyloseq and TreeSE objects

### Load Raw Phyloseq

```{r}
raw_phy <- readr::read_rds(phy_file)
raw_phy
```

### Filter Phyloseq & Add New Metadata

```{r}
# Filter
phy <- phyloseq::subset_samples(raw_phy, SampleID %in% meta_df$SampleID)

# Prepare phy inputs
otu_table <- phyloseq::otu_table(phy)
tax_table <- phyloseq::tax_table(phy)
sam_data <-  
  meta_df %>% 
  dplyr::mutate(SampleID_ = SampleID, .before = 1) %>%
  data.frame(row.names = 1) %>% 
  phyloseq::sample_data() %>% 
  .[phyloseq::sample_names(phy), ]

# Add new metadata to new phyloseq
phy <- phyloseq::phyloseq(otu_table, tax_table, sam_data)
phy
```

### Construct TreeSE Object

```{r}
tse <- mia::convertFromPhyloseq(phy)
tse
```

### Save Phyloseq and TreeSE Objects

```{r}
readr::write_rds(phy, here::here(out_dir, "phy_cleaned.rds"))
readr::write_rds(tse, here::here(out_dir, "tse_cleaned.rds"))
```

## Appendix

### Session Info

```{r}
devtools::session_info()
```

### Contact

- **Analysis Lead:** Francesc Català-Moll
- **Email:** fcatala@irsicaixa.es
- **Institution:** GEM - IrsiCaixa
