---
title: "02 - Alpha Diversity Analysis"
editor_options: 
  chunk_output_type: console
---

## Define the input and output paths

```{r}
# input
tse_file <- here::here("data", "processed", "tse_cleaned.rds")
igc_file <- here::here("data", "raw", "dataTable.rds")

# output
out_dir <- here::here("data", "02_alpha_diversity")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
```

## Set-up

```{r}
# load libraries
library(magrittr)
library(patchwork)
suppressPackageStartupMessages(library(tidySingleCellExperiment))
```

## Computate Alpha Diversity Metrics

### Load TreeSE

```{r}
# load
tse <- readr::read_rds(tse_file)
```

### Add Gene Richness

```{r}
# load
igc_df <- readr::read_rds(igc_file) %>% tibble::as_tibble()

# threshold
threshold <- 
  igc_df %>%
  dplyr::group_by(SampleID) %>% 
  dplyr::summarise(NumberMappedReads = max(NumberMappedReads)) %>% 
  dplyr::summarise(q = quantile(NumberMappedReads, 0.02)) %>%
  dplyr::pull()
  
# extract
igc_df <- 
  igc_df %>% 
  dplyr::filter(ReadCountReal >= threshold) %>% 
  dplyr::group_by(SampleID) %>%
  dplyr::summarise(gene_richness = min(GeneNumber))

# merge
tse <- tse %>% dplyr::left_join(igc_df, by = "SampleID")
```

### Compute Alpha Diversity Metrics

```{r}
# select metrics
metrics <- c(
  "shannon_diversity", "gini_dominance","observed_richness", "pielou_evenness"
)

# compute
tse <-
  tse %>%
  mia::addAlpha(
    index = metrics,
    sample = quantile(colSums(assay(tse, "counts")), 0.02),
    niter = 10,
    BPPARAM = BiocParallel::MulticoreParam()
  )
```

## Plot Alpha Diversity Metrics

```{r}
#| message: FALSE
#| warning: FALSE
plt_list <- 
  c(metrics, "gene_richness") %>% 
  purrr::set_names() %>% 
  purrr::map( ~ {
    # prepare data
    plt_df <- colData(tse) %>% tibble::as_tibble() 
     
    # calculate stats
    stats_2 <-
      plt_df %>%
      rstatix::group_by(treatment) %>%
      rstatix::wilcox_test(
        formula = formula(glue::glue("{.x} ~ time_point")),
        p.adjust.method = "fdr"
      ) %>%
      rstatix::add_significance() %>%
      rstatix::add_xy_position(x = "time_point", group = "treatment") %>% 
      dplyr::mutate(y.position = y.position)
      
    # plot
    plt <- 
      plt_df %>% 
      tidyr::drop_na(!!dplyr::sym(.x)) %>% 
      tidyplots::tidyplot(x = time_point, y = !!dplyr::sym(.x), colour = treatment) %>% 
      tidyplots::add_boxplot(show_outliers = FALSE) %>%
      tidyplots::add_data_points_jitter(alpha = 0.4) %>%
      tidyplots::add_test_asterisks(
        method = "wilcox_test", hide_info = TRUE, bracket.nudge.y = 0.05, tip.length = 0.01
      ) %>%
      tidyplots::add(ggpubr::stat_pvalue_manual(
        stats_2, label = "p.adj.signif", hide.ns = TRUE, tip.length = 0.01,
      )) %>% 
      tidyplots::adjust_x_axis_title("Time Point (weeks)") %>%
      tidyplots::adjust_legend_title("Treatment") %>%
      tidyplots::adjust_y_axis_title(
        .x %>% stringr::str_replace_all("_", " ") %>% stringr::str_to_title()
      ) %>% 
      tidyplots::adjust_colors(tidyplots::colors_discrete_friendly) 
      
    if (.x == "gene_richness") {
      plt <- plt %>% tidyplots::adjust_y_axis(labels = scales::scientific)
    }
    
    # save
    plt %>%
      tidyplots::adjust_size(width = 40, height = 40, unit = "mm") %>%
      tidyplots::save_plot(
        here::here(out_dir, glue::glue("alpha_{.x}.pdf")),
        view_plot = FALSE
      )
    
    plt
  })
```

::: {.panel-tabset}

```{r}
#| results: asis
names(plt_list) %>% 
  purrr::walk(~ {
    cat("#### ", .x, "\n")
    plot(plt_list[[.x]])
    cat("\n\n")
  })
```

:::

## Compute Correlations

### Define Cytokines and Markers

```{r}
populations <- c("CD4", "CD8", "CD4_nadir", "CD4_CD38_DR", "CD8_CD38_DR")
cytokines <- c("CRP", "IL6", "TNFa", "sCD14")
others <- c("HIV_VL", "BMI")

all_markers <- c(populations, cytokines, others)
```

### Cytokines Log2 Transformation

```{r}
tse2 <- tse %>% dplyr::mutate(dplyr::across(dplyr::all_of(all_markers), ~ log2(.x + 1)))
```

### Compute Correlations

```{r}
#| warning: FALSE
corr_df <-
  mia::splitOn(tse2, "treatment") %>%
  purrr::map_dfr(~ {
    mia::getCrossAssociation(
      .x,
      .x,
      col.var1 = c(metrics, "gene_richness"),
      col.var2 = all_markers,
      method = "spearman",
      p_adj_method = "fdr",
      test.signif = TRUE, 
      verbose = FALSE, 
      show_warnings = FALSE
    ) %>%
      dplyr::mutate(treatment = unique(.x$treatment))
  })
```

### Plot Global Correlations

```{r}
#| message: FALSE
#| warning: FALSE
plt_list <- 
  c(metrics, "gene_richness") %>% 
  purrr::set_names() %>% 
  purrr::map(~ {
    # prepare data
    plt_df <-
      corr_df %>% 
      dplyr::filter(Var1 == .x) %>% 
      dplyr::mutate(sign = p_adj < 0.05 & abs(cor) >= 0.2) 
      
    # plot
    .name <- stringr::str_replace(.x, "_", " ") %>% stringr::str_to_title()
    lim <- abs(plt_df$cor) %>% max() %>% round(digits = 1)
    plt <- 
      plt_df %>% 
      ggplot(aes(x = Var2, y = treatment, fill = cor, colour = sign)) +
      geom_point(aes(size = abs(cor), alpha = sign), shape = 22, stroke = 1) +
      scale_fill_gradientn(
        colours = rev(RColorBrewer::brewer.pal(11, "Spectral")),
        limits = c(-lim, lim)
      ) +
      scale_colour_manual(values = c("TRUE" = "black", "FALSE" = "white")) +
      scale_alpha_manual(values = c("TRUE" = 0.9, "FALSE" = 0.6), guide = "none") +
      scale_size_continuous(range = c(2, 8), guide = "none") +
      theme_minimal() +
      labs(
        x = "",
        y = "Treatment", 
        fill = glue::glue("Spearman Correlation\n(Mark. vs. {.name})"), 
        colour = "q < 0.05"
      ) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "top", 
        legend.title.position = "top", 
        legend.title = element_text(hjust = 0.5), 
        legend.key.height = unit(0.5, "cm"),
      ) 
    
    # Resize and save
    plt <- 
      plt + plot_spacer() +
      plot_layout(
        widths = ggplot2::unit(c(70, 1), "mm"), 
        heights = ggplot2::unit(30, "mm")
      ) 
    
    tidyplots::save_plot(
      plt, 
      here::here(out_dir, glue::glue("correlation_{.x}.pdf")), 
      view_plot = FALSE
    )
    
    plt
  })
```

::: {.panel-tabset}

```{r}
#| results: asis
names(plt_list) %>% 
  purrr::walk(~ {
    cat("#### ", .x, "\n")
    plot(plt_list[[.x]])
    cat("\n\n")
  })
```

:::

### Plot Scatter Correlations

```{r}
#| message: FALSE
#| warning: FALSE
plt_df <- colData(tse2) %>% tibble::as_tibble()
plt_list <- 
  c(metrics, "gene_richness") %>% 
  purrr::set_names() %>% 
  purrr::map(function(.metric) {
   all_markers %>% 
      purrr::set_names() %>% 
      purrr::map(function(.marker) {
        # prepare data
        marker_values <- plt_df[[.marker]] %>% .[is.finite(.)]
        lab_min <- min(marker_values, na.rm = TRUE)
        lab_max <- max(marker_values, na.rm = TRUE)
        lab_range <- lab_max - lab_min
        .name <- stringr::str_replace(.metric, "_", " ") %>% stringr::str_to_title()
        plt <- 
          plt_df %>%
          tidyr::drop_na(!!.metric, !!.marker) %>%
          tidyplots::tidyplot(
            x = !!dplyr::sym(.metric),
            y = !!dplyr::sym(.marker),
            colour = treatment
          ) %>%
          tidyplots::add_data_points(alpha = 0.5) %>%
          tidyplots::add(geom_smooth(method = "lm", alpha = 0.1, formula = 'y ~ x')) %>%
          tidyplots::add(ggpubr::stat_cor(
            method = "spearman",
            cor.coef.name = "rho",
            label.y.npc = "bottom",
            p.digits = 1,
            label.y = c(lab_min - 0.05 * lab_range, lab_min - 0.15 * lab_range), 
            size = 3
          )) %>% 
          tidyplots::adjust_legend_title("Treatment") %>%
          tidyplots::adjust_y_axis_title(glue::glue("Log2  {.marker}")) %>% 
          tidyplots::adjust_x_axis_title(.name) %>% 
          tidyplots::adjust_x_axis(rotate_labels = 30) %>% 
          tidyplots::adjust_colors(tidyplots::colors_discrete_friendly) 
          
          if (.metric == "gene_richness") {
            plt <- plt %>% tidyplots::adjust_x_axis(labels = scales::scientific)
          }
      
          # save
          plt %>%
            tidyplots::adjust_size(width = 50, height = 40, unit = "mm") %>%
            tidyplots::save_plot(
              here::here(out_dir, glue::glue("corr_{.metric}_{.marker}.pdf")),
              view_plot = FALSE
            )
          
          plt
      })
  })
```

::: {.panel-tabset}

```{r}
#| results: asis
#| fig.width: 12
#| fig.height: 7
#| warning: FALSE
names(plt_list) %>% 
  purrr::walk(~ {
    cat("#### ", .x, "\n")
    plot(
      plt_list[[.x]] %>% 
        patchwork::wrap_plots() + 
        plot_layout(guides = "collect")
    )
    cat("\n\n")
  })
```

:::

## Appendix

### Session Info

```{r}
devtools::session_info()
```

### Contact

- **Analysis Lead:** Francesc Català-Moll
- **Email:** fcatala@irsicaixa.es
- **Institution:** GEM - IrsiCaixa