---
title: "05 - Differential Abundance Analysis"
---

## Define the input and output paths

```{r}
# input
tse_file <- here::here("data", "processed", "tse_alpha.rds")

# output
out_dir <- here::here("data", "05_diff_abundance")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# confounders 
confounders <- c(
  "treatment",
  "time_point",
  "gender",
  "risk_group",
  "record_id",
  "age", 
  "HIV_VL"
)

# filtering
lfc_cutoff <- 0.5
q_cutoff <- 0.05
prev_cutoff <- 0.2
cor_cutoof <- 0.3
```

## Set-up

```{r}
# load libraries
library(magrittr)
library(patchwork)
library(tidyplots)
suppressPackageStartupMessages(library(tidySingleCellExperiment))
```

## Load TSE Object and Transform

```{r}
tse <- 
  readr::read_rds(tse_file) %>% 
  mia::transformAssay(method = "relab") %>% 
  mia::transformAssay(method = "clr", pseudocount = 1) %>% 
  mia::transformAssay(method = "log10", pseudocount = 1)

tse
```

## Differential Abundance Analysis (ANCOMBC2)

### DA Computation

```{r}
ancombc_results <- ANCOMBC::ancombc2(
  tse,
  fix_formula = "treatment * time_point + gender + age + risk_group + HIV_VL",
  tax_level = "Species",
  p_adj_method = "fdr",
  group = "treatment", 
  pairwise = TRUE, 
  prv_cut = 0.1,
  n_cl = 8L
)
```

### Preprocess Results

```{r}
res_df <- 
  ancombc_results %>% 
  purrr::pluck("res") %>% 
  tibble::as_tibble() %>% 
  dplyr::select(taxon, dplyr::contains("DTG")) %>% 
  tidyr::pivot_longer(
    cols = -taxon,
    names_to   = c("metric", "comparison"),
    names_pattern = "^(lfc|se|W|p|q|diff_robust|diff|passed_ss)_(.+)$",
    values_to  = "value"
  ) %>% 
  dplyr::mutate(comparison = sub("^robust_", "", comparison)) %>% 
  tidyr::pivot_wider(names_from = metric, values_from = value) %>% 
  dplyr::rename(qval = q) %>% 
  dplyr::mutate(
    time_point = case_when(
      comparison == "treatmentDTG" ~ 0,
      stringr::str_ends(comparison, "time_point24") ~ 24, 
      stringr::str_ends(comparison, "time_point48") ~ 48, 
      stringr::str_ends(comparison, "time_point96") ~ 96, 
      TRUE ~ NA_integer_
    ), 
    time_point = factor(time_point), 
    comparison = "DTG vs RAL", 
    .after = 1
  )

# add prevalence
res_df <- 
  assay(tse) %>% 
  tibble::as_tibble(rownames = "taxon") %>%
  tidyr::pivot_longer(cols = -taxon, names_to = "SampleID", values_to = "ab") %>%
  dplyr::group_by(taxon) %>% 
  dplyr::summarise(prevalence = round(sum(ab > 0) / dplyr::n(), 2)) %>% 
  dplyr::left_join(
    rowData(tse) %>% tibble::as_tibble(rownames = "taxon"), 
    by = "taxon"
  ) %>% 
  dplyr::select(taxon, Species, Genus, prevalence) %>% 
  dplyr::left_join(res_df, ., relationship = "many-to-many", by = "taxon")

res_df
```

### Volcano Plot

```{r}
#| message: FALSE
#| warning: FALSE

# prepare data
plot_df <- 
  res_df %>% 
  dplyr::mutate(
    candidate = abs(lfc) >= lfc_cutoff & qval < q_cutoff & passed_ss == 1 & prevalence > prev_cutoff,
    time_point = factor(time_point), 
    neg_log10_qval = -log10(qval)
  )

# prepare plot
tp_palette <- c("#0072B2", "#009E73", "#E69F00", "#D55E00")
max_x_val <- plot_df$lfc %>% abs() %>% max(na.rm = TRUE) * 1.1
max_y_val <- plot_df$neg_log10_qval %>% max(na.rm = TRUE) * 1.1
volcano_plts <- 
  plot_df$time_point %>% 
  levels() %>% 
  purrr::map2(tp_palette, ~ {
    volc_plt <- 
      plot_df %>% 
      dplyr::filter(time_point == .x) %>% 
      tidyr::drop_na(lfc, neg_log10_qval) %>%
      tidyplot(lfc, neg_log10_qval, size = neg_log10_qval) %>% 
      add(geom_point(
        data = filter_rows(!candidate), 
        color = "lightgrey", 
        aes(size = neg_log10_qval)
      )) %>%
      add(geom_point(
        data = filter_rows(candidate), 
        colour = .y, 
        aes(size = neg_log10_qval)
      )) %>%
      add(scale_size_continuous(range = c(0.5, 3), limits = c(0, max_y_val))) %>%
      add_reference_lines(x = c(-0.5, 0.5), y = -log10(0.05)) %>% 
      adjust_x_axis_title("$Log[2]~fold~change$") %>% 
      adjust_y_axis_title("$-Log[10]~italic(P)~adjusted$") %>% 
      adjust_legend_title("Time Point") %>% 
      adjust_x_axis(limits = c(-max_x_val, max_x_val)) %>% 
      adjust_y_axis(limits = c(0, max_y_val))

    # save plot
    volc_plt %>% 
      adjust_size(width = 50, height = 50) %>% 
      save_plot(
        here::here(out_dir, glue::glue("volcano_plot_{.x}.pdf")), 
        view_plot = FALSE
      )

    volc_plt %>% 
      adjust_caption(glue::glue("Time Point: {.x}")) %>% 
      adjust_font(fontsize = 12)

  })

wrap_plots(volcano_plts)
```

### Bar Plot of Top Candidates

```{r}
#| message: FALSE
#| warning: FALSE

# prepare plot
bar_plts <- 
  plot_df$time_point %>% 
  levels() %>% 
  purrr::set_names() %>% 
  purrr::map2(tp_palette, ~ {
    bar_plt <- 
      plot_df %>% 
      dplyr::filter(candidate) %>% 
      dplyr::filter(time_point == .x) %>% 
      dplyr::group_by(Species) %>% 
      # dplyr::filter(abs(lfc) == max(abs(lfc))) %>% 
      dplyr::ungroup() %>%
      dplyr::mutate(
        abs_lfc = abs(lfc), 
        sign = dplyr::if_else(lfc > 0, "Up DTG", "Down DTG"), 
        Species = forcats::fct_reorder(Species, lfc)
      ) %>% 
      tidyplot(abs_lfc, Species, color = sign) %>%
      add_reference_lines(x = c(0.5)) %>% 
      add(geom_col(width = 0.75, alpha = 0.6)) %>% 
      add(geom_text(aes(label = prevalence), hjust = 1.2, size = 3, colour = "black")) %>%
      adjust_y_axis_title(NULL) %>% 
      adjust_x_axis_title("Absolute Log2 Fold Change") %>%
      adjust_colors(c("#FFE577", "#7CC9E5")) 

    # save plot
    bar_plt %>% 
      adjust_size(width = 50, height = 75) %>% 
      save_plot(
        here::here(out_dir, glue::glue("bar_plot_{.x}.pdf")), 
        view_plot = FALSE
      )

    bar_plt %>% 
      adjust_caption(glue::glue("Time Point: {.x}")) %>% 
      adjust_font(fontsize = 10) %>% 
      adjust_size(width = 50, height = 120) 

  })
```

::: panel-tabset
```{r}
#| results: asis
#| warning: FALSE
#| fig.height: 10
names(bar_plts) %>% 
  purrr::walk(~ {
    cat("#### ", .x, "\n")
    plot(
      bar_plts[[.x]]
    )
    cat("\n\n")
  })
```
:::

### Evolution Plot

```{r}
#| message: FALSE
#| warning: FALSE
#| fig.width: 10

plot_df_2 <-
  plot_df %>%
  dplyr::group_by(Species) %>%
  dplyr::filter(any(candidate)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(time_point = as.character(time_point) %>% as.numeric()) 

evolution_plt <- 
  plot_df_2 %>% 
  ggplot(aes(time_point, lfc, colour = Species, group = Species)) +
  geom_line(size = 1, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") + 
  theme_minimal(base_size = 18) +
  theme(legend.position = "none") +
  geom_point(data = plot_df_2 %>% dplyr::filter(candidate), shape = 8, size = 2) +
  ggrepel::geom_text_repel(
    data = dplyr::filter(plot_df_2, time_point == 96),
    aes(label = Species),
    fontface = "bold", size = 3, direction = "y", xlim = c(100, NA), hjust = 0.5,
    segment.size = .7, segment.alpha = .5, segment.linetype = "dotted", box.padding = .2,
    segment.curvature = -0.2, segment.ncp = 3, segment.angle = 20, max.overlaps = 25
  ) +
  coord_cartesian(clip = "off") +
  # scale_colour_manual(values = 
  #   c("#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#66A61E", "#E6AB02", "#A6761D", 
  #     "#666666", "#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00", "#000000", 
  #     "#A65628", "#F781BF", "#999999"
  #   )
  # ) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 140),breaks = c(0, 24, 48, 96)) +
  scale_y_continuous(expand = c(0, 0), breaks = seq(-3, 3, by = 1)) +
  labs(x = "Time Point (weeks)", y = "Log2 Fold Change")
  

# save plot
ggsave(
  plot = evolution_plt,
  filename = here::here(out_dir, "evolution_plot.pdf"), 
  width = 300,
  height = 200,
  units = "mm"
)

evolution_plt
```

### Export Results as Tables

```{r}
# save results
readr::write_csv(plot_df, here::here(out_dir, "diff_abundance_results.csv"))
readr::write_rds(ancombc_results, here::here(out_dir,"ancombc_results.rds"))

# print results
plot_df %>% dplyr::filter(candidate) %>% kableExtra::kable()
```

## Species vs Markers Correlations (Spearman)

### Define Cytokines and Markers

```{r}
populations <- c("CD4", "CD8", "CD4_nadir", "CD4_CD38_DR", "CD8_CD38_DR")
cytokines <- c("CRP", "IL6", "TNFa", "sCD14")

all_markers <- c(populations, cytokines)
```

### Cytokines Log2 Transformation

```{r}
tse2 <- tse %>% dplyr::mutate(dplyr::across(dplyr::all_of(cytokines), ~ log2(.x + 1)))
```

### Compute Correlations

```{r}
#| warning: FALSE
corr_df <-
  mia::getCrossAssociation(
    tse2,
    tse2,
    assay.type1 = "clr",
    col.var2 = all_markers,
    method = "spearman",
    p_adj_method = "fdr",
    test.signif = TRUE,
    verbose = FALSE,
    show_warnings = FALSE
  ) %>% 
  tibble::as_tibble() %>%
    dplyr::mutate(
    Var2 = dplyr::case_when(
      Var2 == "CD4" ~ "CD4+ T cells",
      Var2 == "CD8" ~ "CD8+ T cells",
      Var2 == "CD4_nadir" ~ "CD4+ T cell count nadir",
      Var2 == "CD4_CD38_DR" ~ "CD4+ CD38+ HLA-DR+ T cells",
      Var2 == "CD8_CD38_DR" ~ "CD8+ CD38+ HLA-DR+ T cells",
      Var2 == "IL6" ~ "IL-6",
      Var2 == "TNFa" ~ "TNF-α",
      Var2 == "HIV_VL" ~ "HIV viral load"
    )
  ) 
```

### Plot Best Correlations (Not necessary DA taxa)

```{r}
#| message: FALSE
#| warning: FALSE
#| fig.width: 10

# prepare data
plt_df <- 
  corr_df %>% 
  dplyr::mutate(sign = p_adj < 0.05 & abs(cor) >= cor_cutoof) %>% 
  dplyr::group_by(Var1) %>% 
  dplyr::filter(any(sign)) %>% 
  dplyr::ungroup() %>% 
  dplyr::left_join(
    rowData(tse2) %>% 
      tibble::as_tibble(rownames = "Var1") %>% 
      dplyr::select(Var1, Species)
  )

# plot
lim <- 0.5
plt <-
  plt_df %>%
  ggplot(aes(x = Species, y = Var2, fill = cor, colour = sign)) +
  geom_point(aes(size = abs(cor), alpha = sign), shape = 22, stroke = 1) +
  scale_fill_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "RdYlGn")),
    limits = c(-lim, lim)
  ) +
  scale_colour_manual(values = c("TRUE" = "black", "FALSE" = "white")) +
  scale_alpha_manual(values = c("TRUE" = 0.9, "FALSE" = 0.6), guide = "none") +
  scale_size_continuous(range = c(2, 8), guide = "none") +
  theme_minimal() +
  labs(
    x = "", y = "", fill = "Spearman Correlation", colour = "q < 0.05"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    legend.title.position = "top",
    legend.title = element_text(hjust = 0.5),
    legend.key.height = unit(0.5, "cm")
  )

# Resize and save
r_plt <-
  plt + plot_spacer() +
  plot_layout(
    widths = ggplot2::unit(c(175, 1), "mm"),
    heights = ggplot2::unit(60, "mm")
  )

tidyplots::save_plot(
  r_plt,
  here::here(out_dir, glue::glue("corr_best.pdf")),
  view_plot = FALSE
)

r_plt
```

### Plot Best Correlations for DA taxas

```{r}
#| message: FALSE
#| warning: FALSE

# prepare data
sig_taxa <- plot_df %>% dplyr::filter(candidate) %>% dplyr::pull(taxon)
plt_df <- 
  corr_df %>% 
  dplyr::mutate(sign = p_adj < 0.05) %>% 
  dplyr::filter(Var1 %in% sig_taxa) %>% 
  dplyr::group_by(Var1) %>% 
  dplyr::filter(any(sign)) %>% 
  dplyr::ungroup() %>% 
  dplyr::left_join(
    rowData(tse2) %>% 
      tibble::as_tibble(rownames = "Var1") %>% 
      dplyr::select(Var1, Species)
  )

# plot
lim <- 0.5
plt <-
  plt_df %>%
  ggplot(aes(x = Species, y = Var2, fill = cor, colour = sign)) +
  geom_point(aes(size = abs(cor), alpha = sign), shape = 22, stroke = 1) +
  scale_fill_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "RdYlGn")),
    limits = c(-lim, lim)
  ) +
  scale_colour_manual(values = c("TRUE" = "black", "FALSE" = "white")) +
  scale_alpha_manual(values = c("TRUE" = 0.9, "FALSE" = 0.6), guide = "none") +
  scale_size_continuous(range = c(2, 8), guide = "none") +
  theme_minimal() +
  labs(
    x = "", y = "", fill = "Spearman Correlation", colour = "q < 0.05"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    legend.title.position = "top",
    legend.title = element_text(hjust = 0.5),
    legend.key.height = unit(0.5, "cm")
  )

# Resize and save
r_plt <-
  plt + plot_spacer() +
  plot_layout(
    widths = ggplot2::unit(c(200, 1), "mm"),
    heights = ggplot2::unit(75, "mm")
  )

tidyplots::save_plot(
  r_plt,
  here::here(out_dir, glue::glue("corr_diff_taxa.pdf")),
  view_plot = FALSE
)

r_plt
```

## Compute Repeated Measures Correlation Coefficient Species vs Markers 

### Plot Best Correlations

```{r}
#| message: FALSE
#| warning: FALSE
#| fig.width: 15
#| fig.high: 12

# computes correlations
rmcorr_df_all <-
  tidyr::expand_grid(taxas = rownames(assay(tse2)), all_markers) %>%
  purrr::pmap_dfr(
    ~ {
      # prepare data
      tax_ab <-
        tse2[.x, ] %>%
        assay("clr") %>%
        tibble::as_tibble() %>%
        tidyr::pivot_longer(
          cols = dplyr::everything(),
          names_to = "SampleID",
          values_to = .x
        )

      f_df <-
        colData(tse2) %>%
        tibble::as_tibble() %>%
        dplyr::select(SampleID, dplyr::all_of(c(.y, confounders))) %>%
        dplyr::left_join(tax_ab, by = "SampleID") %>%
        tidyr::drop_na(dplyr::all_of(c(.x, .y))) %>%
        dplyr::mutate(
          dplyr::across(dplyr::where(is.character), as.factor),
          treatment = factor(treatment, levels = c("DRV/r", "DTG"))
        )

      # compute rmcorr
      res <- rmcorr::rmcorr(
        participant = "record_id",
        measure1 = f_df[[.x]],
        measure2 = f_df[[.y]],
        dataset = f_df
      )

      # create tibble
      tibble::tibble(
        taxa = .x,
        marker = .y,
        cor = res$r,
        p = res$p,
        df = list(f_df),
        rmcorr = list(res),
      )
    }
  ) %>%
  dplyr::mutate(p_adj = p.adjust(p, method = "fdr"), .after = p) %>%
  dplyr::arrange(p_adj) %>%
  dplyr::left_join(
    rowData(tse2) %>% tibble::as_tibble(rownames = "taxa"),
    by = "taxa"
  )

# prepare data
plt_df <-
  rmcorr_df_all %>%
  dplyr::mutate(
    sign = p_adj < 0.05 & abs(cor) >= cor_cutoof,
    marker = factor(marker, levels = all_markers),
  ) %>%
  dplyr::group_by(taxa) %>%
  dplyr::filter(any(sign)) %>%
  dplyr::ungroup()

# plot
lim <- abs(plt_df$cor) %>% max(na.rm = T) %>% round(digits = 1) * 1.05
plt <-
  plt_df %>%
  ggplot(aes(x = Species, y = marker, fill = cor, colour = sign)) +
  geom_point(aes(size = abs(cor), alpha = sign), shape = 22, stroke = 1) +
  scale_fill_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "RdYlGn")),
    limits = c(-lim, lim)
  ) +
  scale_colour_manual(values = c("TRUE" = "black", "FALSE" = "white")) +
  scale_alpha_manual(values = c("TRUE" = 0.9, "FALSE" = 0.6), guide = "none") +
  scale_size_continuous(range = c(2, 8), guide = "none") +
  theme_minimal() +
  labs(
    x = "",
    y = "",
    fill = "Repeated Measures Correlation",
    colour = "q < 0.05"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    legend.title.position = "top",
    legend.title = element_text(hjust = 0.5),
    legend.key.height = unit(0.5, "cm"),
  )

# Resize and save
r_plt <-
  plt +
  plot_spacer() +
  plot_layout(
    widths = ggplot2::unit(c(300, 1), "mm"),
    heights = ggplot2::unit(100, "mm")
  )

tidyplots::save_plot(
  r_plt,
  here::here(out_dir, glue::glue("rm_top_correlation.pdf")),
  view_plot = FALSE
)

r_plt
```

### Plot Correlations for DA Taxas

```{r}
#| message: FALSE
#| warning: FALSE
#| fig.width: 10

# prepare data
plt_df <-
  rmcorr_df_all %>%
  dplyr::filter(taxa %in% sig_taxa) %>%
  dplyr::mutate(
    sign = p_adj < 0.05,
    marker = factor(marker, levels = all_markers),
  ) %>%
  dplyr::group_by(taxa) %>%
  dplyr::filter(any(sign)) %>%
  dplyr::ungroup()

# plot
lim <- abs(plt_df$cor) %>% max(na.rm = TRUE) %>% round(digits = 1) * 1.05
plt <-
  plt_df %>%
  ggplot(aes(x = Species, y = marker, fill = cor, colour = sign)) +
  geom_point(aes(size = abs(cor), alpha = sign), shape = 22, stroke = 1) +
  scale_fill_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "RdYlGn")),
    limits = c(-lim, lim)
  ) +
  scale_colour_manual(values = c("TRUE" = "black", "FALSE" = "white")) +
  scale_alpha_manual(values = c("TRUE" = 0.9, "FALSE" = 0.6), guide = "none") +
  scale_size_continuous(range = c(2, 8), guide = "none") +
  theme_minimal() +
  labs(
    x = "",
    y = "",
    fill = "Repeated Measures Correlation",
    colour = "q < 0.05"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    legend.title.position = "top",
    legend.title = element_text(hjust = 0.5),
    legend.key.height = unit(0.5, "cm"),
  )

# Resize and save
r_plt <-
  plt +
  plot_spacer() +
  plot_layout(
    widths = ggplot2::unit(c(150, 1), "mm"),
    heights = ggplot2::unit(75, "mm")
  )

tidyplots::save_plot(
  r_plt,
  here::here(out_dir, glue::glue("rm_DA_correlation.pdf")),
  view_plot = FALSE
)

r_plt
```

## Appendix

### Session Info

```{r}
devtools::session_info()
```

### Contact

- **Analysis Lead:** Francesc Català-Moll
- **Email:** fcatala@irsicaixa.es
- **Institution:** GEM - IrsiCaixa
