---
title: "05 - Differential Abundance Analysis"
---

## Define the input and output paths

```{r}
# input
tse_file <- here::here("data", "processed", "tse_alpha.rds")

# output
out_dir <- here::here("data", "05_diff_abundance")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# confounders 
confounders <- c(
  "treatment",
  "time_point",
  "gender",
  "risk_group",
  "record_id",
  "age"
)

prev_cutoff <- 0.10
```

## Set-up

```{r}
# load libraries
library(magrittr)
library(patchwork)
library(tidyplots)
suppressPackageStartupMessages(library(tidySingleCellExperiment))
```

## Load TSE Object and Transform

```{r}
tse <- 
  readr::read_rds(tse_file) %>% 
  mia::transformAssay(method = "relab") %>% 
  mia::transformAssay(method = "clr", pseudocount = 1) %>% 
  mia::transformAssay(method = "log10", pseudocount = 1) 

tse
```

## Differential Abundance Analysis (ANCOMBC2)

### DA Computation

```{r}
ancombc_results <- ANCOMBC::ancombc2(
  tse,
  fix_formula = "treatment * time_point + gender + age + risk_group",
  rand_formula = "(1 | record_id)", 
  tax_level = "Species",
  p_adj_method = "fdr",
  group = "treatment", 
  pairwise = TRUE, 
  prv_cut = 0.1,
  n_cl = 8L
)
```

### Preprocess Results

```{r}
res_df <- 
  ancombc_results %>% 
  purrr::pluck("res") %>% 
  tibble::as_tibble() %>% 
  dplyr::select(taxon, dplyr::contains("DTG")) %>% 
  tidyr::pivot_longer(
    cols = -taxon,
    names_to   = c("metric", "comparison"),
    names_pattern = "^(lfc|se|W|p|q|diff_robust|diff|passed_ss)_(.+)$",
    values_to  = "value"
  ) %>% 
  dplyr::mutate(comparison = sub("^robust_", "", comparison)) %>% 
  tidyr::pivot_wider(names_from = metric, values_from = value) %>% 
  dplyr::rename(qval = q) %>% 
  dplyr::mutate(
    time_point = case_when(
      comparison == "treatmentDTG" ~ 0,
      stringr::str_ends(comparison, "time_point24") ~ 24, 
      stringr::str_ends(comparison, "time_point48") ~ 48, 
      stringr::str_ends(comparison, "time_point96") ~ 96, 
      TRUE ~ NA_integer_
    ), 
    time_point = factor(time_point), 
    comparison = "DTG vs RAL", 
    .after = 1
  )

# add prevalence
res_df <- 
  assay(tse) %>% 
  tibble::as_tibble(rownames = "taxon") %>%
  tidyr::pivot_longer(cols = -taxon, names_to = "SampleID", values_to = "ab") %>%
  dplyr::group_by(taxon) %>% 
  dplyr::summarise(prevalence = round(sum(ab > 0) / dplyr::n(), 2)) %>% 
  dplyr::left_join(
    rowData(tse) %>% tibble::as_tibble(rownames = "taxon"), 
    by = "taxon"
  ) %>% 
  dplyr::select(taxon, Species, Genus, prevalence) %>% 
  dplyr::left_join(res_df, ., relationship = "many-to-many", by = "taxon")

# add ci
res_df <- 
  res_df %>% 
  dplyr::mutate(
    ci_low  = lfc - 1.96 * se,
    ci_high = lfc + 1.96 * se
  )

res_df
```

## Forest plot for best taxas

positive estimates ⇒ DTG > DRVr
negative estimates ⇒ DTG < DRVr

```{r}
# get top taxas
top_taxa <- 
  res_df %>% 
  dplyr::filter((ci_low > 0 | ci_high < 0) & prevalence > prev_cutoff) %>% 
  dplyr::slice_min(qval, n = 20) %>% 
  dplyr::pull(Species)

top_taxa_ids <- 
  res_df %>% 
  dplyr::filter((ci_low > 0 | ci_high < 0) & prevalence > prev_cutoff) %>% 
  dplyr::slice_min(qval, n = 20) %>% 
  dplyr::pull(taxon)

order <- 
  res_df %>% 
  dplyr::filter(Species %in% top_taxa & time_point == "96") %>% 
  dplyr::mutate(Species = forcats::fct_reorder(Species, lfc)) %>% 
  dplyr::pull(Species) %>% 
  levels()

forest_plt <- 
  res_df %>%
  dplyr::filter(Species %in% top_taxa) %>% 
  dplyr::mutate(
    Species = factor(Species, levels = order),
    ci95_excludes_zero = ci_low > 0 | ci_high < 0, 
    ci95_label = ifelse(ci95_excludes_zero, "Yes", "No")
  ) %>% 
  tidyplots::tidyplot(x = lfc, y = Species, colour = ci95_label) %>%
  tidyplots::add(geom_pointrange(
    aes(xmin = ci_low, xmax = ci_high),
    alpha = 0.8,
    position = position_dodge(width = 0.6),
    size = 2 / ggplot2::.pt
  )) %>%
  tidyplots::adjust_colors(c("#737373", "#887DE6")) %>% 
  tidyplots::add_reference_lines(x = 0) %>% 
  tidyplots::adjust_font(fontsize = 12) %>% 
  tidyplots::adjust_x_axis_title("") %>% 
  tidyplots::adjust_y_axis_title(NULL) %>% 
  tidyplots::adjust_legend_title("95% CI excludes 0") %>% 
  tidyplots::adjust_legend_position("bottom") %>% 
  tidyplots::adjust_size(250, 100) %>% 
  tidyplots::add(facet_wrap(~time_point, nrow = 1)) %>% 
  tidyplots::save_plot(here::here(out_dir, "forest_plt.pdf"), view_plot = FALSE)


forest_plt
```

## Effect Heatamap

```{r}
# prepare matrix
tp_levels <- levels(res_df$time_point)
mat <-
  res_df %>%
  dplyr::filter(Species %in% top_taxa) %>%
  dplyr::select(Species, time_point, lfc) %>%
  tidyr::pivot_wider(names_from = time_point, values_from = lfc) %>%
  tibble::column_to_rownames("Species") %>%
  as.matrix() %>%
  .[order[order %in% rownames(.)], tp_levels, drop = FALSE] %>% 
  .[rev(rownames(.)), , drop = FALSE]

# Ranges and Colours
rng <- max(abs(mat), na.rm = TRUE)
if (!is.finite(rng) || rng == 0) { rng <- 1 }
col_fun <- circlize::colorRamp2(
  c(-rng, 0, rng), 
  c("#2b8cbe", "#f7f7f7", "#d7301f")
)

# Prevalence annotation
prev_df <-
  res_df %>%
  dplyr::distinct(Species, prevalence) %>%
  dplyr::filter(Species %in% rownames(mat)) %>%
  dplyr::arrange(match(Species, rownames(mat)))

row_anno <- ComplexHeatmap::rowAnnotation(
  Prevalence = ComplexHeatmap::anno_barplot(
    prev_df$prevalence,
    gp = grid::gpar(fill = "#9ecae1"),
    width = unit(18, "mm")
  ),
  annotation_label = "Prevalence"
)

# Heatmap
ht <- ComplexHeatmap::Heatmap(
  mat,
  name = "logFC\n(DTG vs RAL)",
  col = col_fun,
  na_col = "grey90",
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_order = rownames(mat),
  column_title = "Time point (hours)",
  column_split = factor(colnames(mat), levels = tp_levels),
  border = TRUE,
  row_names_gp = grid::gpar(fontsize = 9),
  column_names_gp = grid::gpar(fontsize = 10),
  heatmap_width = unit(140, "mm"),
  heatmap_height = unit(180, "mm"),
  show_row_dend = FALSE,
  show_column_dend = FALSE
)

# plot
Cairo::CairoPDF(here::here(out_dir, "heatmap.pdf"), width = 210/25.4, height = 297/25.4)
ComplexHeatmap::draw(
  row_anno + ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right",
  merge_legend = TRUE
)
dev.off()

ComplexHeatmap::draw(
  row_anno + ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right",
  merge_legend = TRUE
)
```


### Export Results as Tables

```{r}
# save results
readr::write_csv(res_df, here::here(out_dir, "diff_abundance_results.csv"))
readr::write_rds(ancombc_results, here::here(out_dir,"ancombc_results.rds"))
```

## Species vs Markers Correlations (Spearman)

### Define Cytokines and Markers

```{r}
populations <- c("CD4", "CD8", "CD4_CD38_DR", "CD8_CD38_DR")
cytokines <- c("CRP", "IL6", "TNFa", "sCD14")
other <- "BMI"

all_markers <- c(populations, cytokines, other)

all_markers_long <- c(
  "CD4+ T cells",
  "CD8+ T cells",
  "CD4+ CD38+ HLA-DR+ T cells",
  "CD8+ CD38+ HLA-DR+ T cells",
  "CRP",
  "IL-6",
  "TNF-a",
  "sCD14",
  "BMI"
)
```

### Cytokines Log2 Transformation

```{r}
tse2 <- 
  tse %>% 
  dplyr::mutate(dplyr::across(dplyr::all_of(cytokines), ~ log2(.x + 1)))
```

## Compute Repeated Measures Correlation Coefficient Species vs Markers 

```{r}
#| warning: FALSE
rmcorr_df <-
  tse2[top_taxa_ids, ] %>% 
  mia::splitOn("treatment") %>%
  purrr::map_dfr(function(.split) {
    tidyr::expand_grid(taxas = rownames(assay(.split)), all_markers) %>%
      purrr::pmap_dfr(~ {
        tax_ab <-
          .split[.x, ] %>%
          assay("clr") %>%
          tibble::as_tibble() %>%
          tidyr::pivot_longer(
            cols = dplyr::everything(),
            names_to = "SampleID",
            values_to = .x
          )

      f_df <-
        colData(.split) %>%
        tibble::as_tibble() %>%
        dplyr::select(SampleID, dplyr::all_of(c(.y, confounders))) %>%
        dplyr::left_join(tax_ab, by = "SampleID") %>%
        tidyr::drop_na(dplyr::all_of(c(.x, .y))) %>%
        dplyr::mutate(
          dplyr::across(dplyr::where(is.character), as.factor),
          treatment = factor(treatment, levels = c("DRV/r", "DTG"))
        )

      # compute rmcorr
      res <- rmcorr::rmcorr(
        participant = "record_id",
        measure1 = f_df[[.x]],
        measure2 = f_df[[.y]],
        dataset = f_df
      )

      # create tibble
      tibble::tibble(
        taxa = .x,
        marker = .y,
        treatment = f_df$treatment %>% unique(), 
        cor = res$r,
        p = res$p,
        df = list(f_df),
        rmcorr = list(res), 

      )
    })
  }) %>% 
  dplyr::group_by(treatment, marker) %>%
  dplyr::mutate(p_adj = p.adjust(p, method = "fdr")) %>%
  dplyr::ungroup() 
```

## Plot

```{r}
plt_df <-
  rmcorr_df %>%
  dplyr::left_join(rowData(tse2) %>% tibble::as_tibble(rownames = "taxa")) %>% 
  dplyr::mutate(
    marker = dplyr::case_when(
      marker == "CD4" ~ "CD4+ T cells", 
      marker == "CD8" ~ "CD8+ T cells", 
      marker == "CD4_CD38_DR" ~ "CD4+ CD38+ HLA-DR+ T cells", 
      marker == "CD8_CD38_DR" ~ "CD8+ CD38+ HLA-DR+ T cells", 
      marker == "IL6" ~ "IL-6", 
      marker == "TNFa" ~ "TNF-a", 
      TRUE ~ marker
    ),
    sign = dplyr::case_when(
      p_adj < 0.05 ~ "< 0.05",
      TRUE ~ "ns"
    ),
    marker = factor(marker, levels = all_markers_long), 
    Species = factor(Species, levels = order)
  )

# plot
plt <-
  plt_df %>%
  ggplot(aes(x = marker, y = Species, fill = cor, colour = sign)) +
  geom_point(aes(size = abs(cor), alpha = sign), shape = 22, stroke = 1) +
  scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11, "PiYG")), limits = c(-0.6, 0.6), breaks = c(-0.6, 0, 0.6)) +
  scale_colour_manual(values = c("< 0.05" = "black", "ns" = "white")) +
  scale_alpha_manual(values = c("< 0.05" = 0.9, "ns" = 0.6), guide = "none") +
  scale_size_continuous(range = c(3, 10), guide = "none") +
  theme_minimal(base_size = 15) +
  labs(
    x = "",
    y = "",
    fill = glue::glue("Repeated Measures Correlation"),
    colour = "q < 0.05"
  ) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  facet_wrap(~ treatment, nrow = 1) 

# Resize and save
plt <-
  plt +
  plot_spacer() +
  plot_layout(
    widths = ggplot2::unit(c(200, 1), "mm"),
    heights = ggplot2::unit(150, "mm")
  )

tidyplots::save_plot(
  plt,
  here::here(out_dir, glue::glue("rm_correlation_treatment_all.pdf")),
  view_plot = FALSE
)

plt
```


## Appendix

### Session Info

```{r}
devtools::session_info()
```

### Contact

- **Analysis Lead:** Francesc Català-Moll
- **Email:** fcatala@irsicaixa.es
- **Institution:** GEM - IrsiCaixa
