---
title: "06 - Functional Analysis (Path Abundance)"
---

## Define the input and output paths

```{r}
# input
tse_file <- here::here("data", "processed", "tse_alpha.rds")
meta_file <- here::here("data", "processed", "metadata_cleaned.csv")
abund_file <- here::here("data", "raw", "All_pathabun.tsv")

# output
out_dir <- here::here("data", "06_functional_analysis")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
```

## Set-up

```{r}
# load libraries
library(magrittr)
library(patchwork)
library(tidyplots)
suppressPackageStartupMessages(library(tidySingleCellExperiment))
```

## Construct TSE Object

```{r}
# load data
tse <- readr::read_rds(tse_file)

# filter and formatig path abun
path_ab <- 
  readr::read_tsv(abund_file, show_col_types = FALSE) %>% 
  dplyr::filter(!stringr::str_detect(`# Pathway`, "[|]")) %>% 
  dplyr::rename_with(~ stringr::str_remove_all(.x, "_Abundance")) %>%
  dplyr::select(1, dplyr::all_of(tse$SampleID)) %>% 
  data.frame(row.names = 1) %>% 
  as.matrix()

# prepare rowdata
row_dat <- 
  readr::read_tsv(abund_file, show_col_types = FALSE) %>% 
  dplyr::filter(!stringr::str_detect(`# Pathway`, "[|]")) %>% 
  dplyr::select(pathway_ = `# Pathway`) %>%
  dplyr::mutate(Species = pathway_) %>% 
  data.frame() %>% 
  as.matrix()

# make tse
tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(counts = path_ab), 
  colData = colData(tse), 
  rowData = row_dat
)
```

## Differential Abundance Analysis (ANCOMBC2)

### DA Computation

```{r}
if (file.exists(here::here(out_dir, "ancombc_results.rds"))) {
  ancombc_results <- readr::read_rds(here::here(out_dir, "ancombc_results.rds"))
} else {
  ancombc_results <- ANCOMBC::ancombc2(
    tse,
    fix_formula = "treatment * time_point + gender + age + aids_event + risk_group + HIV_VL",
    tax_level = "Species",
    p_adj_method = "fdr",
    group = "treatment",
    prv_cut = 0.1,
    n_cl = 8
  )
}
```

### Preprocess Results

```{r}
pre_df <- ancombc_results %>% purrr::pluck("res") %>% tibble::as_tibble()

# lfc_df
lfc_df <- 
  pre_df %>% 
  dplyr::select(taxon, dplyr::starts_with("lfc_treatmentDTG")) %>% 
  dplyr::rename_with(.cols = dplyr::starts_with("lfc"), .fn = ~ {
    stringr::str_remove_all(.x, ".*point") %>% 
      stringr::str_replace_all("lfc_treatmentDTG", "0")
  }) %>% 
  dplyr::mutate(`24` = `24` + `0`, `48` = `48` + `0`, `96` = `96` + `0`) %>% 
  tidyr::pivot_longer(cols = -taxon, names_to = "time_point", values_to = "lfc") 

# pval_df
pval_df <- 
  pre_df %>% 
  dplyr::select(taxon, dplyr::starts_with("p_treatmentDTG")) %>% 
  dplyr::rename_with(.cols = dplyr::starts_with("p_"), .fn = ~ {
    stringr::str_remove_all(.x, ".*point") %>% 
      stringr::str_replace_all("p_treatmentDTG", "0")
  }) %>% 
  tidyr::pivot_longer(cols = -taxon, names_to = "time_point", values_to = "pval")

# qval_df
qval_df <- 
  pre_df %>% 
  dplyr::select(taxon, dplyr::starts_with("q_treatmentDTG")) %>% 
  dplyr::rename_with(.cols = dplyr::starts_with("q_"), .fn = ~ {
    stringr::str_remove_all(.x, ".*point") %>% 
      stringr::str_replace_all("q_treatmentDTG", "0")
  }) %>% 
  tidyr::pivot_longer(cols = -taxon, names_to = "time_point", values_to = "qval")

# passed DTG
passed_df <- 
  pre_df %>% 
  dplyr::select(taxon, dplyr::starts_with("passed_ss_treatmentDTG")) %>% 
  dplyr::rename_with(.cols = dplyr::starts_with("p"), .fn = ~ {
    stringr::str_remove_all(.x, ".*point") %>% 
      stringr::str_replace_all("passed_ss_treatmentDTG", "0")
  }) %>% 
  tidyr::pivot_longer(cols = -taxon, names_to = "time_point", values_to = "passed")

# prepare annot
annot_df <- 
  rowData(tse) %>%
  tibble::as_tibble(rownames = "taxon") %>% 
  dplyr::select(taxon, Species)

# combine
res_df <- 
  lfc_df %>% 
  dplyr::left_join(pval_df, by = c("taxon", "time_point")) %>% 
  dplyr::left_join(qval_df, by = c("taxon", "time_point")) %>% 
  dplyr::left_join(passed_df, by = c("taxon", "time_point")) %>% 
  dplyr::left_join(annot_df, by = "taxon")

# add prevalence
res_df <- 
  assay(tse) %>% 
  tibble::as_tibble(rownames = "taxon") %>%
  tidyr::pivot_longer(cols = -taxon, names_to = "SampleID", values_to = "ab") %>%
  dplyr::group_by(taxon) %>% 
  dplyr::summarise(prevalence = round(sum(ab > 0) / dplyr::n(), 2)) %>% 
  dplyr::left_join(res_df, ., by = "taxon")
```

### Volcano Plot

```{r}
# prepare data
plot_df <- 
  res_df %>% 
  dplyr::mutate(
    candidate = abs(lfc) >= 0.5 & qval < 0.05 & passed & prevalence > 0.1 & time_point != "0",
    time_point = factor(time_point), 
    neg_log10_qval = -log10(qval)
  )

# prepare plot
volc_plt <- 
  plot_df %>% 
  tidyr::drop_na(lfc, neg_log10_qval) %>%
  tidyplot(lfc, neg_log10_qval, color = time_point, size = neg_log10_qval) %>% 
  add(geom_point(
    data = filter_rows(!candidate), 
    color = "lightgrey", 
    aes(size = neg_log10_qval)
  )) %>%
  add(geom_point(data = filter_rows(candidate), aes(size = neg_log10_qval))) %>%
  add(scale_size_continuous(range = c(0.5, 3))) %>%
  add_reference_lines(x = c(-0.5, 0.5), y = -log10(0.05)) %>% 
  adjust_x_axis_title("$Log[2]~fold~change$") %>% 
  adjust_y_axis_title("$-Log[10]~italic(P)~adjusted$") %>% 
  adjust_legend_title("Time Point") %>% 
  adjust_size(width = 50, height = 50)

# save plot
save_plot(volc_plt, here::here(out_dir, "volcano_plot.pdf"), view_plot = FALSE)

volc_plt
```

### Bar Plot of Top Candidates

```{r}
# prepare plot
bar_plt <- 
  plot_df %>% 
  dplyr::filter(candidate) %>% 
  dplyr::group_by(Species) %>% 
  # dplyr::filter(abs(lfc) == max(abs(lfc))) %>% 
  dplyr::ungroup() %>%
  dplyr::mutate(
    abs_lfc = abs(lfc), 
    sign = dplyr::if_else(lfc > 0, "Up DTG", "Down DTG"), 
    Species = stringr::str_c(Species, "_", time_point, "w"), 
    Species = forcats::fct_reorder(Species, lfc)
  ) %>% 
  tidyplot(abs_lfc, Species, color = sign) %>%
  add_reference_lines(x = c(0.5)) %>% 
  add(geom_col(width = 0.75, alpha = 0.6)) %>% 
  add(geom_text(aes(label = prevalence), hjust = 1.2, size = 3, colour = "black")) %>%
  adjust_y_axis_title(NULL) %>% 
  adjust_x_axis_title("Absolute Log2 Fold Change") %>%
  adjust_colors(c("#FFE577", "#7CC9E5")) %>%
  adjust_size(width = 75, height = 75)

# save plot
save_plot(bar_plt, here::here(out_dir, "barplot_plot.pdf"), view_plot = FALSE)

bar_plt
```

### Lolliplot Of Top Candidates

```{r}
lolli_plt <- 
  plot_df %>% 
  dplyr::filter(candidate) %>% 
  dplyr::group_by(Species) %>% 
  # dplyr::filter(abs(lfc) == max(abs(lfc))) %>% 
  dplyr::ungroup() %>%
  dplyr::mutate(
    abs_lfc = abs(lfc), 
    sign = dplyr::if_else(lfc > 0, "Up DTG", "Down DTG"), 
    Species = stringr::str_c(Species, "_", time_point, "w"), 
    Species = forcats::fct_reorder(Species, lfc)
  ) %>% 
  tidyplot(abs_lfc, Species, color = sign, size = prevalence) %>% 
  add(geom_point(alpha = 1)) %>% 
  add_reference_lines(x = c(0.5)) %>% 
  add_mean_bar(width = 0.2, alpha = 0.7) %>% 
  adjust_size(width = 75, height = 75) %>% 
  adjust_x_axis_title("Absolute Log2 Fold Change") %>%
  adjust_colors(c("Down DTG" = "#FFE577", "Up DTG" = "#7CC9E5")) %>%
  add(scale_size_continuous(range = c(4, 8), breaks = c(0.2, 0.4, 0.6, 0.8))) %>%
  adjust_y_axis_title(NULL)  %>%
  add(geom_text(aes(label = prevalence), hjust = -1, size = 3)) %>%
  add(scale_x_continuous(limits = c(0, 2.4))) %>%
  adjust_size(width = 75, height = 75)

# save plot
save_plot(lolli_plt, here::here(out_dir, "lolli_plot.pdf"), view_plot = FALSE)

lolli_plt
```

```{r}
plot_df_2 <-
  plot_df %>%
  dplyr::group_by(Species) %>%
  dplyr::filter(any(candidate)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(time_point = as.character(time_point) %>% as.numeric()) 

evolution_plt <- 
  plot_df_2 %>% 
  ggplot(aes(time_point, lfc, colour = Species, group = Species)) +
  geom_line(size = 1, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") + 
  theme_minimal() +
  theme(legend.position = "none") +
  geom_point(data = plot_df_2 %>% dplyr::filter(candidate), shape = 8, size = 2) +
  ggrepel::geom_text_repel(
    data = dplyr::filter(plot_df_2, time_point == 96),
    aes(label = Species),
    fontface = "bold", size = 3, direction = "y", xlim = c(100, NA), hjust = 0.5,
    segment.size = .7, segment.alpha = .5, segment.linetype = "dotted", box.padding = .2,
    segment.curvature = -0.2, segment.ncp = 3, segment.angle = 20
  ) +
  coord_cartesian(clip = "off", ylim = c(-1, 2)) +
  scale_colour_manual(values = 
    c("#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#66A61E", "#E6AB02", "#A6761D", 
      "#666666", "#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00", "#000000", 
      "#A65628", "#F781BF", "#999999"
    )
  ) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 180),breaks = c(0, 24, 48, 96)) +
  scale_y_continuous(expand = c(0, 0), breaks = seq(-3, 3, by = 1)) +
  labs(x = "Time Point (weeks)", y = "Log2 Fold Change")
  
# save plot
ggsave(
  plot = evolution_plt,
  filename = here::here(out_dir, "evolution_plot.pdf"), 
  width = 275,
  height = 100,
  units = "mm"
)

evolution_plt
```


### Export Results as Tables

```{r}
# save results
readr::write_csv(plot_df, here::here(out_dir, "diff_abundance_results.csv"))
readr::write_rds(ancombc_results, here::here(out_dir,"ancombc_results.rds"))

# print results
plot_df %>% dplyr::filter(candidate) %>% kableExtra::kable()
```

## Correlations Sign Taxa vs Markers

### Define Cytokines and Markers

```{r}
populations <- c("CD4", "CD8", "CD4_nadir", "CD4_CD38_DR", "CD8_CD38_DR")
cytokines <- c("CRP", "IL6", "TNFa", "sCD14")

all_markers <- c(populations, cytokines)
```

### Cytokines Log2 Transformation

```{r}
tse2 <- 
  tse %>% 
  dplyr::mutate(dplyr::across(dplyr::all_of(cytokines), ~ log2(.x + 1))) %>% 
  mia::transformAssay(method = "clr", pseudocount = 1)
```

### Compute Correlations

```{r}
#| warning: FALSE
corr_df <-
  mia::getCrossAssociation(
    tse2,
    tse2,
    assay.type1 = "clr",
    col.var2 = all_markers,
    method = "spearman",
    p_adj_method = "fdr",
    test.signif = TRUE,
    verbose = FALSE,
    show_warnings = FALSE
  ) %>% 
  tibble::as_tibble()
```

### Plot Correlations 

```{r}
#| message: FALSE
#| warning: FALSE

# prepare data
sig_taxa <- plot_df %>% dplyr::filter(candidate) %>% dplyr::pull(taxon)
plt_df <- 
  corr_df %>% 
  dplyr::filter(Var1 %in% sig_taxa) %>% 
  dplyr::mutate(sign = p_adj < 0.05 & abs(cor) >= 0.1) 

# plot
lim <- abs(plt_df$cor) %>% max() %>% round(digits = 1)
lim <- 0.3
plt <-
  plt_df %>%
  ggplot(aes(x = Var1, y = Var2, fill = cor, colour = sign)) +
  geom_point(aes(size = abs(cor), alpha = sign), shape = 22, stroke = 1) +
  scale_fill_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "PiYG")),
    limits = c(-lim, lim)
  ) +
  scale_colour_manual(values = c("TRUE" = "black", "FALSE" = "white")) +
  scale_alpha_manual(values = c("TRUE" = 0.9, "FALSE" = 0.6), guide = "none") +
  scale_size_continuous(range = c(2, 8), guide = "none") +
  theme_minimal() +
  labs(
    x = "", y = "", fill = "Spearman Correlation", colour = "q < 0.05"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    legend.title.position = "top",
    legend.title = element_text(hjust = 0.5),
    legend.key.height = unit(0.5, "cm")
  )

# Resize and save
r_plt <-
  plt + plot_spacer() +
  plot_layout(
    widths = ggplot2::unit(c(120, 1), "mm"),
    heights = ggplot2::unit(60, "mm")
  )

tidyplots::save_plot(
  r_plt,
  here::here(out_dir, glue::glue("corr_diff_functions.pdf")),
  view_plot = FALSE
)

r_plt
```

## Appendix

### Session Info

```{r}
devtools::session_info()
```

### Contact

- **Analysis Lead:** Francesc Català-Moll
- **Email:** fcatala@irsicaixa.es
- **Institution:** GEM - IrsiCaixa
