---
title: "07 - Functional Analysis (pfam)"
---

## Define the input and output paths

```{r}
# input
tse_file <- here::here("data", "processed", "tse_alpha.rds")
fun_file <- here::here("data", "raw", "All_genefam.tsv")

# output
out_dir <- here::here("data", "07_functional_pfam")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
```

## Set-up

```{r}
# load libraries
library(magrittr)
library(patchwork)
library(tidyplots)
suppressPackageStartupMessages(library(tidySingleCellExperiment))
```

## Construct TSE Object

```{r eval=FALSE}
# load metadata
m_data <- 
  readr::read_rds(tse_file) %>% 
  colData(tse) %>% 
  tibble::as_tibble(rownames = "xx") %>% 
  data.frame(row.names = 1)

# filter and formatig gen fam
tbl <- 
  data.table::fread(fun_file, sep = "\t") %>%
  tibble::as_tibble() %>% 
  dplyr::rename_with(~ stringr::str_remove_all(.x, "_Abundance.*")) %>% 
  dplyr::rename("function_id" = 1) %>% 
  dplyr::filter(!stringr::str_detect(function_id, "[|]")) %>% 
  dplyr::select(function_id, rownames(m_data)) %>% 
  data.frame(row.names = 1)

# SE
humann_se <- SummarizedExperiment::SummarizedExperiment(
  assays = list(humann = tbl),
  colData = m_data
)

# regroup
humann_se <- microfunk::humann_regroup(humann_se, to = "pfam")

readr::write_rds(humann_se, here::here(out_dir, "humann_se.rds"))
```

```{r}
humann_se <- readr::read_rds(here::here(out_dir, "humann_se.rds"))
```

## Differential Abundance Analysis (ANCOMBC2)

### DA Computation

```{r eval=FALSE}
ancombc_results <- ANCOMBC::ancombc2(
  assay(humann_se, "humann"),
  meta_data = colData(humann_se),
  fix_formula = "treatment * time_point + gender + age + aids_event + risk_group + HIV_VL",
  p_adj_method = "fdr",
  group = "treatment",
  prv_cut = 0.1,
  n_cl = 8
)
```

```{r}
ancombc_results <- readr::read_rds(here::here(out_dir, "ancombc_results.rds"))
```

### Preprocess Results

```{r}
pre_df <- ancombc_results %>% purrr::pluck("res") %>% tibble::as_tibble()

# lfc_df
lfc_df <- 
  pre_df %>% 
  dplyr::select(taxon, dplyr::starts_with("lfc_treatmentDTG")) %>% 
  dplyr::rename_with(.cols = dplyr::starts_with("lfc"), .fn = ~ {
    stringr::str_remove_all(.x, ".*point") %>% 
      stringr::str_replace_all("lfc_treatmentDTG", "0")
  }) %>% 
  dplyr::mutate(`24` = `24` + `0`, `48` = `48` + `0`, `96` = `96` + `0`) %>% 
  tidyr::pivot_longer(cols = -taxon, names_to = "time_point", values_to = "lfc") 

# pval_df
pval_df <- 
  pre_df %>% 
  dplyr::select(taxon, dplyr::starts_with("p_treatmentDTG")) %>% 
  dplyr::rename_with(.cols = dplyr::starts_with("p_"), .fn = ~ {
    stringr::str_remove_all(.x, ".*point") %>% 
      stringr::str_replace_all("p_treatmentDTG", "0")
  }) %>% 
  tidyr::pivot_longer(cols = -taxon, names_to = "time_point", values_to = "pval")

# qval_df
qval_df <- 
  pre_df %>% 
  dplyr::select(taxon, dplyr::starts_with("q_treatmentDTG")) %>% 
  dplyr::rename_with(.cols = dplyr::starts_with("q_"), .fn = ~ {
    stringr::str_remove_all(.x, ".*point") %>% 
      stringr::str_replace_all("q_treatmentDTG", "0")
  }) %>% 
  tidyr::pivot_longer(cols = -taxon, names_to = "time_point", values_to = "qval")

# passed DTG
passed_df <- 
  pre_df %>% 
  dplyr::select(taxon, dplyr::starts_with("passed_ss_treatmentDTG")) %>% 
  dplyr::rename_with(.cols = dplyr::starts_with("p"), .fn = ~ {
    stringr::str_remove_all(.x, ".*point") %>% 
      stringr::str_replace_all("passed_ss_treatmentDTG", "0")
  }) %>% 
  tidyr::pivot_longer(cols = -taxon, names_to = "time_point", values_to = "passed")

# prepare annot
annot_df <- 
  rowData(humann_se) %>%
  tibble::as_tibble(rownames = "taxon") %>% 
  dplyr::select(taxon, id_name)

# combine
res_df <- 
  lfc_df %>% 
  dplyr::left_join(pval_df, by = c("taxon", "time_point")) %>% 
  dplyr::left_join(qval_df, by = c("taxon", "time_point")) %>% 
  dplyr::left_join(passed_df, by = c("taxon", "time_point")) %>% 
  dplyr::left_join(annot_df, by = "taxon")

# add prevalence
res_df <- 
  assay(humann_se) %>% 
  tibble::as_tibble(rownames = "taxon") %>%
  tidyr::pivot_longer(cols = -taxon, names_to = "SampleID", values_to = "ab") %>%
  dplyr::group_by(taxon) %>% 
  dplyr::summarise(prevalence = round(sum(ab > 0) / dplyr::n(), 2)) %>% 
  dplyr::left_join(res_df, ., by = "taxon")
```

### Volcano Plot

```{r}
# prepare data
plot_df <- 
  res_df %>%
  dplyr::filter(!stringr::str_detect(id_name, "unknown function|Uncharacterized")) %>% 
  dplyr::mutate(
    candidate = abs(lfc) >= 1 & qval < 0.05 & passed & prevalence > 0.1 & time_point != "0",
    time_point = factor(time_point), 
    neg_log10_qval = -log10(qval)
  )

# prepare plot
volc_plt <- 
  plot_df %>% 
  tidyr::drop_na(lfc, neg_log10_qval) %>%
  tidyplot(lfc, neg_log10_qval, color = time_point, size = neg_log10_qval) %>% 
  add(geom_point(
    data = filter_rows(!candidate), 
    color = "lightgrey", 
    aes(size = neg_log10_qval)
  )) %>%
  add(geom_point(data = filter_rows(candidate), aes(size = neg_log10_qval))) %>%
  add(scale_size_continuous(range = c(0.5, 3))) %>%
  add_reference_lines(x = c(-1, 1), y = -log10(0.05)) %>% 
  adjust_x_axis_title("$Log[2]~fold~change$") %>% 
  adjust_y_axis_title("$-Log[10]~italic(P)~adjusted$") %>% 
  adjust_legend_title("Time Point") %>% 
  adjust_size(width = 50, height = 50)

# save plot
save_plot(volc_plt, here::here(out_dir, "volcano_plot.pdf"), view_plot = FALSE)

volc_plt
```

### Bar Plot of Top Candidates

```{r}
# prepare plot
bar_plt <- 
  plot_df %>% 
  dplyr::filter(candidate) %>% 
  dplyr::group_by(id_name) %>% 
  dplyr::filter(abs(lfc) == max(abs(lfc))) %>% 
  dplyr::ungroup() %>%
  dplyr::mutate(
    abs_lfc = abs(lfc), 
    sign = dplyr::if_else(lfc > 0, "Up DTG", "Down DTG"), 
    id_name = forcats::fct_reorder(id_name, lfc)
  ) %>% 
  tidyplot(abs_lfc, id_name, color = sign) %>%
  add_reference_lines(x = 1) %>% 
  add(geom_col(width = 0.75, alpha = 0.6)) %>% 
  add(geom_text(aes(label = prevalence), hjust = 1.2, size = 3, colour = "black")) %>%
  adjust_y_axis_title(NULL) %>% 
  adjust_x_axis_title("Absolute Log2 Fold Change") %>%
  adjust_colors(c("#FFE577", "#7CC9E5")) %>%
  adjust_size(width = 75, height = 150)

# save plot
save_plot(bar_plt, here::here(out_dir, "barplot_plot.pdf"), view_plot = FALSE)

bar_plt
```

### Export Results as Tables

```{r}
# save results
readr::write_csv(plot_df, here::here(out_dir, "diff_abundance_results.csv"))
readr::write_rds(ancombc_results, here::here(out_dir, "ancombc_results.rds"))

# print results
plot_df %>% dplyr::filter(candidate) %>% kableExtra::kable()
```

## Correlations Sign Taxa vs Markers

### Define Cytokines and Markers

```{r}
populations <- c("CD4", "CD8", "CD4_nadir", "CD4_CD38_DR", "CD8_CD38_DR")
cytokines <- c("CRP", "IL6", "TNFa", "sCD14")

all_markers <- c(populations, cytokines)
```

### Cytokines Log2 Transformation

```{r}
tse2 <- 
  TreeSummarizedExperiment::TreeSummarizedExperiment(
    list(counts = assay(humann_se)), 
    colData = colData(humann_se), 
    rowData = rowData(humann_se)
  ) %>% 
  dplyr::mutate(dplyr::across(dplyr::all_of(cytokines), ~ log2(.x + 1))) %>% 
  mia::transformAssay(method = "clr", pseudocount = 1)
```

### Compute Correlations

```{r}
#| warning: FALSE
corr_df <-
  mia::getCrossAssociation(
    tse2,
    tse2,
    assay.type1 = "clr",
    col.var2 = all_markers,
    method = "spearman",
    p_adj_method = "fdr",
    test.signif = TRUE,
    verbose = FALSE,
    show_warnings = FALSE
  ) %>% 
  tibble::as_tibble()
```

### Plot Correlations 

```{r}
#| message: FALSE
#| warning: FALSE

# prepare data
sig_taxa <- plot_df %>% dplyr::filter(candidate) %>% dplyr::pull(taxon)
plt_df <- 
  corr_df %>% 
  dplyr::filter(Var1 %in% sig_taxa) %>% 
  dplyr::left_join(rowData(tse2) %>% tibble::as_tibble(rownames = "Var1"), by = "Var1") %>% 
  dplyr::mutate(sign = p_adj < 0.05 & abs(cor) >= 0.1) 

# plot
lim <- abs(plt_df$cor) %>% max() %>% round(digits = 1)
lim <- 0.25
plt <-
  plt_df %>%
  ggplot(aes(x = id_name, y = Var2, fill = cor, colour = sign)) +
  geom_point(aes(size = abs(cor), alpha = sign), shape = 22, stroke = 1) +
  scale_fill_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "PiYG")),
    limits = c(-lim, lim)
  ) +
  scale_colour_manual(values = c("TRUE" = "black", "FALSE" = "white")) +
  scale_alpha_manual(values = c("TRUE" = 0.9, "FALSE" = 0.6), guide = "none") +
  scale_size_continuous(range = c(2, 8), guide = "none") +
  theme_minimal() +
  labs(
    x = "", y = "", fill = "Spearman Correlation", colour = "q < 0.05"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    legend.title.position = "top",
    legend.title = element_text(hjust = 0.5),
    legend.key.height = unit(0.5, "cm")
  )

# Resize and save
r_plt <-
  plt + plot_spacer() +
  plot_layout(
    widths = ggplot2::unit(c(200, 1), "mm"),
    heights = ggplot2::unit(120, "mm")
  )

tidyplots::save_plot(
  r_plt,
  here::here(out_dir, glue::glue("corr_diff_functions.pdf")),
  view_plot = FALSE
)

r_plt
```

## Appendix

### Session Info

```{r}
devtools::session_info()
```

### Contact

- **Analysis Lead:** Francesc Català-Moll
- **Email:** fcatala@irsicaixa.es
- **Institution:** GEM - IrsiCaixa

