---
title: "Untitled"
format: html
---

## Define the input and output paths

```{r}
# output
out_dir <- here::here("data", "11_metahiv_integration")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# input
metadata_file <- here::here(out_dir, "merged_metadata.rds")
s3_object <- "s3://mistral-wp6-advanz4/metagenome/WMGS/taxprofiler_metahiv/metaphlan/metaphlan_db1_combined_reports.txt"
```

## Set-Up

```{r}
# load libraries
library(magrittr)
library(patchwork)
suppressPackageStartupMessages(library(tidySingleCellExperiment))
```

## Load Data and transform to tse

### Load metadata

```{r}
metadata_df <- readr::read_rds(metadata_file)
metadata_df
```

### Load Metaphlan table

```{r}
mpa_df <- aws.s3::s3read_using(
  readr::read_tsv, 
  object = s3_object, 
  skip = 1, 
  show_col_types = FALSE
)

mpa_df
```

### Prepare assay table

```{r}
assay_mtx <- 
  mpa_df %>% 
  dplyr::filter(stringr::str_detect(clade_name, "t__")) %>% 
  dplyr::rename_all(~ stringr::str_remove_all(.x, "_metahiv.*|_advanz.*")) %>% 
  dplyr::mutate(clade_name = stringr::str_remove_all(clade_name, ".*[|]")) %>%
  dplyr::select(clade_name, dplyr::all_of(metadata_df$SampleID)) %>% 
  data.frame(row.names = 1) %>% 
  as.matrix()
```

### Prepare rowData

```{r}
it <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "Taxon")
rowdata <- 
  mpa_df %>% 
  dplyr::select(clade_name) %>% 
  dplyr::filter(stringr::str_detect(clade_name, "t__")) %>% 
  tidyr::separate(clade_name, into = it, sep = "\\|") %>%
  dplyr::mutate(id = Taxon, .before = 1) %>%
  data.frame(row.names = 1) 
```

### TSE construction

```{r}
tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
  assays = list(relab = assay_mtx), 
  rowData = rowdata,
  colData = metadata_df
)

tse <- tse %>% mia::agglomerateByRank(rank = "Species")
tse
```

## Dissimilarity Analysis 

### Compute

```{r}
tse <- 
  tse %>% 
  mia::addDissimilarity(assay.type = "relab" , method = "bray") %>% 
  dplyr::mutate(Profile = dplyr::case_when(
    time_point == "0" ~ Profile,
    time_point != "0" ~ paste0(Profile, "_", time_point), 
    TRUE ~ NA_character_
    )
  )

bray_mat <- metadata(tse)[["bray"]] %>% as.matrix()
meta <- as.data.frame(colData(tse))
idx_neg <- which(meta$Profile == "negative")
meta$dist_to_negative <- vapply(
  seq_len(nrow(meta)),
  function(i) mean(bray_mat[i, idx_neg], na.rm = TRUE),
  numeric(1)
)
```

### Plot dissimilarity distributions

```{r}
#| fig-height: 7

profile_cols <- c(
  "DRV/r"               = "#0072B2",
  "DRV/r_24"            = "#3390C3",
  "DRV/r_48"            = "#66AED4",
  "DRV/r_96"            = "#99CCE5",
  "DTG"                 = "#D55E00",
  "DTG_24"              = "#E27F33",
  "DTG_48"              = "#EFA066",
  "DTG_96"              = "#FBC199",
  "concordant"          = "#CC79A7",
  "discordant"          = "#E8A6CA",
  "early_treated"       = "#F0E442",
  "elite_controllers"   = "#F7E985",
  "late_presenters"     = "#F18E6B",
  "naive"               = "#B9CF6A",
  "negative"            = "#999999",
  "viremic_controllers" = "#C2B0E2"
)

plt <- 
  meta %>% 
  dplyr::mutate(
    Profile = forcats::fct_reorder(Profile, dist_to_negative, .fun = median),
    type = dplyr::case_when(
      stringr::str_starts(Profile, "DTG") ~ "DTG", 
      stringr::str_starts(Profile, "DRV/r") ~ "DRV/r", 
      TRUE ~ "metaHIV"
    ),
    type = factor(type, levels = c("metaHIV", "DRV/r", "DTG"))
  ) %>% 
  tidyplots::tidyplot(x = Profile, y = dist_to_negative, colour = Profile) %>% 
  tidyplots::add_boxplot(show_outliers = FALSE) %>%
  tidyplots::add_data_points_jitter(alpha = 0.4, jitter_width = 0.6) %>% 
  tidyplots::adjust_x_axis(rotate_labels = TRUE) %>% 
  tidyplots::adjust_x_axis_title(NULL) %>%
  tidyplots::remove_legend() %>% 
  tidyplots::adjust_y_axis_title("Mean Bray-Curtis dissimilarity to HIV-negative samples") %>% 
  tidyplots::adjust_colors(profile_cols) %>% 
  tidyplots::adjust_font(fontsize = 12) %>% 
  tidyplots::add(facet_wrap(~type, nrow = 1, scales = "free_x", space = "free_x")) %>% 
  tidyplots::adjust_size(width = 125, height = 75, unit = "mm")
      
# save
plt %>%
  tidyplots::save_plot(
    here::here(out_dir, glue::glue("dissimilarity_metahiv.pdf")),
    view_plot = FALSE
  )

plt 
```

### Test differences to negative with lmm

```{r}
# prepare data
plot_df <- 
  meta %>% 
  lme4::lmer(dist_to_negative ~ Profile + (1 | record_id), data = .) %>% 
  emmeans::emmeans(~ Profile) %>% 
  emmeans::contrast(method = "trt.vs.ctrl", ref = "negative") %>% 
  broom.mixed::tidy(conf.int = TRUE) %>% 
  rstatix::add_significance(p.col = "adj.p.value") %>% 
  dplyr::mutate(
    contrast = stringr::str_remove_all(contrast, " .*|[(]|[)]"), 
    contrast = forcats::fct_reorder(contrast, estimate)
  ) 

# plot
plt <- 
  plot_df %>% 
  dplyr::mutate(
    type = dplyr::case_when(
      stringr::str_starts(contrast, "DTG") ~ "DTG", 
      stringr::str_starts(contrast, "DRV/r") ~ "DRV/r", 
      TRUE ~ "metaHIV"
    ),
    type = factor(type, levels = rev(c("metaHIV", "DRV/r", "DTG")))
  ) %>% 
  tidyplots::tidyplot(x = estimate, y = contrast) %>%
  tidyplots::add(geom_pointrange(
    aes(xmin = conf.low, xmax = conf.high),
    alpha = 0.8,
    position = position_dodge(width = 0.6),
    size = 2 / ggplot2::.pt,
    colour = "#252525"
  )) %>%
  tidyplots::add_reference_lines(x = 0) %>%
  tidyplots::adjust_x_axis_title("Estimated difference in Bray-Curtis dissimilarity vs HIV-negative") %>%
  tidyplots::adjust_y_axis_title(NULL) %>% 
  tidyplots::add(
    facet_wrap(~type, ncol = 1, scales = "free_y", space = "free_y", strip.position = "left")
  ) %>% 
  tidyplots::adjust_font(fontsize = 12) %>% 
  tidyplots::adjust_size(width = 75, height = 125, unit = "mm")

# save
plt %>%
  tidyplots::save_plot(
    here::here(out_dir, glue::glue("dissimilarity_forest_metahiv.pdf")),
    view_plot = FALSE
  )

plt
```

## NMDS Analysis 

### Commpute NMDS 

```{r}
tse <- mia::runNMDS(
  tse,
  assay.type = "relab",
  method = "bray",
  nmds.fun = "monoMDS",
  keep.dist = TRUE,
  ntop = 100
)
```

### Scatter Plot (NMDS1 vs NMDS2)

```{r}
#| fig-height: 7

plot_df <- 
  tse %>% 
  dplyr::select(HIV_Status, Profile, study, time_point, dplyr::contains("NMD")) 

# main scatterplot (NMDS1 vs NMDS2)
plt <- 
  plot_df %>% 
  tidyplots::tidyplot(x = NMDS1, y = NMDS2, colour = Profile, linetype = time_point) %>% 
  tidyplots::add_data_points(size = 2) %>% 
  tidyplots::add_ellipse(type = "norm") %>% 
  tidyplots::adjust_size(100, 100) %>% 
  tidyplots::adjust_font(fontsize = 10) %>% 
  tidyplots::adjust_colors(profile_cols)


# save
plt %>%
  tidyplots::save_plot(
    here::here(out_dir, glue::glue("nmds_scater_metahiv.pdf")),
    view_plot = FALSE
  )

plt
```

## Centroid Distances 

### Computation 

```{r}
# negative centorid
neg_centroid <- 
  tse %>%
  dplyr::select(SampleID, HIV_Status, dplyr::contains("NMD")) %>% 
  dplyr::filter(HIV_Status == "negative") %>% 
  dplyr::summarise(
    neg_NMDS1 = mean(NMDS1, na.rm = TRUE),
    neg_NMDS2 = mean(NMDS2, na.rm = TRUE)
  )

# compute distances to negative centorid
tse <- 
  tse %>%
  dplyr::mutate(
    dist_to_negative_centroid = sqrt(
      (NMDS1 - neg_centroid$neg_NMDS1)^2 +
      (NMDS2 - neg_centroid$neg_NMDS2)^2
    )
  ) 
```

### Plot centroid distances to negatives

```{r}
#| fig-height: 7

# plot
plt <- 
  colData(tse) %>%
  tibble::as_tibble() %>% 
  dplyr::mutate(
    Profile = forcats::fct_reorder(Profile, dist_to_negative_centroid, .fun = median, na.rm = TRUE)
  ) %>% 
  tidyplots::tidyplot(x = Profile, y = dist_to_negative_centroid, colour = Profile) %>% 
  tidyplots::add_boxplot(show_outliers = FALSE) %>%
  tidyplots::add_data_points_jitter(alpha = 0.4) %>%
  tidyplots::adjust_size(width = 100, height = 100, unit = "mm") %>% 
  tidyplots::adjust_x_axis(rotate_labels = TRUE) %>% 
  tidyplots::adjust_x_axis_title(NULL) %>%
  tidyplots::remove_legend() %>% 
  tidyplots::adjust_y_axis_title("Distance to HIV1 Negative Centroid") %>% 
  tidyplots::adjust_colors(profile_cols) %>% 
  tidyplots::adjust_font(fontsize = 12)
      
# save
plt %>%
  tidyplots::save_plot(
    here::here(out_dir, glue::glue("centroid_distance_metahiv.pdf")),
    view_plot = FALSE
  )

plt 
```

### Test differences to negative with lmm

```{r}
# prepare data
plot_df <- 
  colData(tse) %>%
  tibble::as_tibble() %>% 
  lme4::lmer(dist_to_negative_centroid ~ Profile + (1 | record_id), data = .) %>% 
  emmeans::emmeans(~ Profile) %>% 
  emmeans::contrast(method = "trt.vs.ctrl", ref = "negative") %>% 
  broom.mixed::tidy(conf.int = TRUE) %>% 
  rstatix::add_significance(p.col = "adj.p.value") %>% 
  dplyr::mutate(
    contrast = stringr::str_remove_all(contrast, " .*|[(]|[)]"), 
    contrast = forcats::fct_reorder(contrast, estimate)
  ) 

# plot
plt <- 
  plot_df %>% 
  tidyplots::tidyplot(x = estimate, y = contrast) %>%
  tidyplots::add(geom_pointrange(
    aes(xmin = conf.low, xmax = conf.high),
    alpha = 0.8,
    position = position_dodge(width = 0.6),
    size = 2 / ggplot2::.pt,
    colour = "#252525"
  )) %>%
  tidyplots::add_reference_lines(x = 0) %>%
  tidyplots::adjust_x_axis_title(
    "Estimated difference in distance to HIV-negative centroid"
  ) %>%
  tidyplots::adjust_y_axis_title(NULL) 

# save
plt %>%
  tidyplots::adjust_size(width = 60, height = 50, unit = "mm") %>% 
  tidyplots::save_plot(
    here::here(out_dir, glue::glue("dist_centroid_forest_metahiv.pdf")),
    view_plot = FALSE
  )

plt %>% 
  tidyplots::adjust_size(width = 75, height = 75, unit = "mm") %>% 
  tidyplots::adjust_font(fontsize = 10)
```

## Appendix

### Session Info

```{r}
devtools::session_info()
```

### Contact

- **Analysis Lead:** Francesc Català-Moll
- **Email:** fcatala@irsicaixa.es
- **Institution:** GEM - IrsiCaixa
